---
title: "IC-read"
author: "Martin Schobben" 
output: 
  rmarkdown::html_vignette:
    fig_caption: true
    pandoc_args:  [
      "--csl", "~/Documents/work/projects/Glendonite_project/common/aquatic-microbial-ecology.csl",
      "--citation-abbreviations", "~/Documents/work/projects/Glendonite_project/common/abbreviations.json", "--filter", "~/.cabal/bin/pandoc-crossref"
                  ]
bibliography: biblio.bib
vignette: >
  %\VignetteIndexEntry{IC-read}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Reading Raw Ion Count Data and Metadata

Raw count data was extracted and collated into a single data frame from files with the extension *"is_txt"* found in the directories associated with the SIMS measurements on glendonite specimens and a belemnite reference, which had been labelled with the suffix *"-GLENDON"*. 

The function `ext_SIMS()`, which takes a single argument; the directory file name, was written to collect the data and transform it into a single file containing; the file name (`file`), the counts (`Y`) and associated measurement order (`X`). The data is complemented with metadata of the associated measurement; the number of measurements (`n`), the magnetic field flux (`B`), the radius (`R`), the mass of measured species (`M`), the assigned name of the species (`species`), the measurement time in milliseconds (`Tc (ms)`), the assigned sample name (`sample`), and the date and time of the measurement (`date`).



```{r setup}
#library(point)
devtools::load_all(".")
```

